---
title: "Analysis Steps"
output: html_document
author: Yousuf Abdelfatah
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

```{r packages, include = FALSE}
# load packages

library(tidyverse)
library(sf)
library(tmap)
library(corrplot)

```

```{r data, include = FALSE}
# load data

detention <- 
  read_csv("Data/Full_Detention_Data.csv")

egypt <- 
  st_read("Data/egy_admbnda_adm1_capmas_20170421/egy.shp")

egypt_population <- 
  read_csv('Data/egypt_population_estimate_compas_2012_v1.1_0_fis.csv') %>% 
  select(ADM1_EN, 'Population' = Total)

socioeconomic <- 
  read_csv("Data/socio_economic.csv")
```

This project conducts a geospatial analysis of politically motivated
arbitrary detention in Egypt. The github repository for this project is
available here:
<https://github.com/yousufabdelfatah/GIS-Arbitrary-Detention>

# Data Acquisition

Data for this project was collected from three sources. Socioeconomic
data was sourced from the Global Data Lab's
[Wealth](https://globaldatalab.org/areadata/2017/wealth-poverty-assets/EGY/?levels=1%2B2%2B3%2B5%2B4&interpolation=1&extrapolation=1&extrapolation_years=3&nearest_real=0)
and [Subnational Human
Development](https://globaldatalab.org/areadata/2017/subnational-human-development/EGY/?levels=1%2B2%2B3%2B5%2B4&interpolation=1&extrapolation=1&extrapolation_years=3&nearest_real=0)
datasets.

Data on detention came from an associated partner human rights
organization with a focus on case documentation. That data was compiled
based on case files and testimony collected by researchers and human
rights lawyers.

The Egypt country [shape
file](https://data.humdata.org/dataset/egypt-administrative-boundaries-levels-0-3)
and [population
estimates](https://data.humdata.org/dataset/egypt-administrative-level-1-governorate-population-statistics/resource/1d8f8cbd-68ec-4149-a26b-5ef10a4cae99)
originate from the Egyptian Central Agency for Public Mobilization and
Statistics. They were hosted and downloaded from the Humanitarian Data
Exchange.

# Data Cleaning

Data Cleaning and Data Analysis were conducted in separate scripts in
order to keep the workflows separate and to avoid accidentally
overwriting the clean data.

The original dataset was split across three excel workbooks and divided
up by individual case numbers. Each case was additionally broken out
into worksheets focused on specific categories such as detention
conditions, legal representation, torture, and enforced disappearances.
Additionally, each case has a "profiles" page containing demographic
information on the individual as well as the charges being leveled
against them

The first step of the data cleaning process was to pull the relevant
sheets from each workbook. Because each workbook was structured
differently I did this through indexing and using the sheet number as
opposed to the sheet name to avoid the troubleshooting required when
working with Arabic text in R Studio. I then iterated through all of the
sheets using the purrr package's map function to make all column names
lower case. Next I renamed misspelled variables and dropped variables I
deemed to provide little value to the analysis. Finally, I standardized
the data types across similar columns.

After I felt comfortable that the values and features had been properly
standardized I used row bind to connect sheets that covered similar
topics and then dropped observations for which all values were missing.
I then joined these new dataframes together to create one large
detention dataframe and united redundant columns.

The final dataset consisted of 420 observations with 64 features (21 of
which were columns outlining the various charges). This I then wrote to
a CVS to be used in the analysis script.

The socioeconomic indicators were much easier to clean. They were
already properly formatted and didn't require much manipulation. I
joined the Wealth and Human Development datasets together and wrote them
to a csv as well for use in the analysis script.

# Data Analysis

I started the data analysis by inspecting the number of detainees from
each governate. To do so I had to standardize the Arabic spellings of
the names of the governate across each dataset and then join the
detention data with the Egypt shape file.

```{r governates, include = FALSE}
# there are some typos in the governates in the detention data
# so standardize those

typo_bs <- c("بنى سويف\n" ,"بنى سويف", "بني سويف") 
correct_bs <- 'بنى سويف'

typo_alex <- "الاسكندرية\n"
correct_alex <- "الإسكندرية"

typo_bh <- "البحيرة\n"
correct_bh <- 'البحيرة'

typo_dum <- "دمياط\n"
correct_dum <- "دمياط"

typo_ism <- "الإسماعيلية\n"
correct_ism <- "الإسماعيلية"

typo_ks <- "كفر الشيخ\n"
correct_ks <- "كفر الشيخ"

typo_qub <- "القليوبية\n"
correct_qub <- 'القليوبية'

typo_min <- "المنيا\n"
correct_min <- "المنيا"

typo_mun <- "المنوفية\n"
correct_mun <- "المنوفية"

typo_ss <- "شمال سيناء\n"
correct_ss <- "شمال سيناء"

typo_sharq <- "الشرقية\n"
correct_sharq <- "الشرقية"

# standardize governate names across datasets

egypt <- 
  egypt %>% 
  mutate(ADM1_AR = case_when(
    ADM1_AR %in% typo_bs ~ correct_bs,
    ADM1_AR %in% typo_alex ~ correct_alex,
    ADM1_AR %in% typo_bh ~ correct_bh,
    ADM1_AR %in% typo_dum ~ correct_dum,
    ADM1_AR %in% typo_ism ~ correct_ism,
    ADM1_AR %in% typo_ks ~ correct_ks,
    ADM1_AR %in% typo_qub ~ correct_qub,
    ADM1_AR %in% typo_min ~ correct_min,
    ADM1_AR %in% typo_mun ~ correct_mun,
    ADM1_AR %in% typo_ss ~ correct_ss,
    ADM1_AR %in% typo_sharq ~ correct_sharq,
    TRUE ~ ADM1_AR))

detention <-
  detention %>% 
  mutate(governate = case_when(
    governate %in% typo_bs ~ correct_bs,
    governate %in% typo_alex ~ correct_alex,
    governate %in% typo_bh ~ correct_bh,
    governate %in% typo_dum ~ correct_dum,
    governate %in% typo_ism ~ correct_ism,
    governate %in% typo_ks ~ correct_ks,
    governate %in% typo_qub ~ correct_qub,
    governate %in% typo_min ~ correct_min,
    governate %in% typo_mun ~ correct_mun,
    governate %in% typo_ss ~ correct_ss,
    governate %in% typo_sharq ~ correct_sharq,
    TRUE ~ governate))

# join shapefile with detention data

joined_data <-
  egypt %>% 
  left_join(detention, 
            by = c("ADM1_AR" = "governate"))

sf_use_s2(FALSE)
```

To visualize the results I made use of the tmap package and also joined
the population data in order to display both the absolute number of
detainees and the number adjusted for the various population size of
each governate.

```{r map_detentions}
# interactive tmap

tmap_mode('view')

# enclosing the map in brackets allows me to implement leaflet widgets to 
# hide layers

{joined_data %>%
    #as_tibble() %>% 
    group_by(ADM1_REF) %>% 
    summarise('Total Detainees' = n()) %>% 
    
    tm_shape(name = 'total detention') +
    tm_polygons(col = 'Total Detainees') +
    tm_basemap('Esri.WorldTopoMap') +
    tm_basemap('Esri.WorldStreetMap') +
    
    joined_data %>%
    group_by(ADM1_EN) %>% 
    left_join(egypt_population, by = 'ADM1_EN') %>% 
    mutate(n = n(),
           detention_per_million = (n/Population)*1E6) %>% 
    
    tm_shape(name = 'detention per million') +
    tm_polygons(col = 'detention_per_million')} %>% 
  
  tmap_leaflet() %>%
  leaflet::hideGroup('detention per million')
```

Next I inspected the distribution of ages of detainees.

```{r age}
# inspect distribution of ages

joined_data %>% 
  ggplot(aes(age)) +
  geom_histogram(aes(y =..density..),
                 bins=25,
                 color="black", 
                 fill="dodgerblue",
                 alpha = 0.8) +
  labs(title="Distribution of Detainee Ages",
       x= "Age")+
  geom_density(color = "orange", size = 1.5) +
  ggthemes::theme_few() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

Then the methods of torture detainees were subjected to.

```{r torture}
# Plot methods and incidence of torture

joined_data %>% 
  as_tibble() %>% 
  select(beating, 
         electrocution, 
         threaten_family, 
         hanging) %>% 
  replace(is.na(.), 0) %>% 
  summarise(beating = sum(beating), 
            electrocution = sum(electrocution), 
            threaten_family = sum(threaten_family),
            hanging = sum(hanging)) %>% 
  rename_with(str_to_title) %>% 
  rename("Threaten Family" = Threaten_family) %>% 
  pivot_longer(everything(), 
               names_to = "Method",
               values_to = "Incidence") %>% 
  ggplot(aes(reorder(Method, -Incidence), Incidence)) +
  geom_col(fill = "red") +
  coord_flip() +
  labs(title="Subjection to Torture",
       x= "Method")+
  theme_test() +
  theme(plot.title = element_text(hjust = 0.5)) 

```

Then the reported conditions of the detention facilities.

```{r conditions}
#inspect detention conditions

joined_data %>% 
  as_tibble() %>% 
  select(42:51) %>% 
  mutate(across(everything(), as.character)) %>% 
  replace(!is.na(.), '1') %>% 
  replace(is.na(.), '0') %>% 
  mutate(across(everything(), as.numeric)) %>% 
  rename_with(str_to_title) %>% 
  summarise_all(sum) %>% 
  rename("Poor Treatment" = Poor_treatment,
         "Poor or No Ventilation" = Ventilation,
         "Solitary Confinement" = Solitary,
         "Denial of Education" = Education,
         "Witholding Food" = Food,
         "No Exercie" = Exercise,
         "Medical Negligence" = Health,
         "Denial of Visitation" = Visitation) %>%
  select(-Strike) %>% 
  pivot_longer(everything(), 
               names_to = "Condition",
               values_to = "Incidence") %>% 
  ggplot(aes(reorder(Condition, -Incidence), Incidence)) +
  geom_col(fill = "green") +
  coord_flip() +
  labs(title="Inhumane Conditions",
       x= "Conditions")+
  theme_test() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

Next I analyzed whether or not detainees were able to access legal
representation. To do so I had to define an English value for the
variable indicating the presence of a lawyer and mutate the Arabic to
correspond with the English value.

```{r lawyer}
# Define English variable for presence of lawyer

Present <- "حضور"

# Plot presence of lawyer

joined_data %>% 
  as_tibble() %>% 
  select(lawyer_present) %>% 
  drop_na()  %>% 
  mutate(
    lawyer_present = case_when(
      lawyer_present == Present ~ "Lawyer Present",
      lawyer_present != Present ~ "No Legal Representation")) %>% 
  ggplot(aes(lawyer_present)) +
  geom_bar(fill = "blue") +
  labs(title="Access to Legal Representation",
       x= "Was a Lawyer Present?")+
  theme_test() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

The last due process abuse I analyzed was whether the detainee was
subjected to enforced disappearance. To do so I defined a variable
indicating whether or not a detainee was disappeared based on whether or
not they had reported the number of days disappeared.

```{r disappeared}

# Define variable for disappeared

joined_data$disappeared <-
  if_else(is.na(joined_data$days_disappeared) == FALSE, 1, 0)

# Plot disappeared

joined_data %>% 
  as_tibble() %>% 
  select(official_arrest:type_of_location, disappeared) %>% 
  mutate(disappeared = if_else(
    disappeared == 1, "Disappearance Reported", "No Report")) %>% 
  ggplot(aes(as.factor(disappeared))) +
  geom_bar(fill = "purple") +
  labs(title="Enforced Disappearances",
       x= "Disappeared")+
  theme_test() +
  theme(plot.title = element_text(hjust = 0.5)) 

joined_data %>% 
  as_tibble() %>% 
  ggplot(aes(days_disappeared)) +
  geom_histogram(fill = "pink", alpha = 0.5, bins = 10) +
  labs(title = "Days Disappeared",
       x = "Days",
       y = "# of Detainees") +
  theme_test() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

In addition to inspecting total instances of abuse, I analyzed torture,
enforced disappearance, and legal representation by governate. I created
static maps for all three of these.

```{r static_map}
basemap <-
  st_bbox(joined_data) %>% 
  tmaptools::read_osm()

tmap_mode('plot')

torture_map <- tm_shape(basemap) +
  tm_rgb() +
  
  joined_data %>% 
  mutate(torture = if_else(
    is.na(beating) == FALSE | 
      is.na(electrocution) == FALSE |
      is.na(threaten_family) == FALSE |
      is.na(hanging) == FALSE,
    1,
    0)) %>% 
  group_by(ADM1_EN) %>% 
  summarise('Reported Torture' = sum(torture)) %>% 
  
  tm_shape(name = 'Reported Torture') +
  tm_polygons(col = 'Reported Torture') +
  
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

rep_map <- tm_shape(basemap) +
  tm_rgb() + 
  
  joined_data %>% 
  mutate(
    lawyer_present = case_when(
      lawyer_present == Present ~ 0,
      lawyer_present != Present ~ 1)) %>% 
  select(lawyer_present, ADM1_EN) %>% 
  drop_na() %>% 
  group_by(ADM1_EN) %>% 
  summarise("Access to Lawyer" = sum(lawyer_present)) %>% 
  
  tm_shape(name = "Access to Lawyer") +
  tm_polygons(col = "Access to Lawyer") +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")

disappearance_map <- tm_shape(basemap) +
  tm_rgb() + 
  
  joined_data %>% 
  select(disappeared, ADM1_EN) %>% 
  group_by(ADM1_EN) %>% 
  summarise(Disappeared = sum(disappeared)) %>% 
  tm_shape(name = "Disappeared") +
  tm_polygons(col = "Disappeared") +
  tm_layout(legend.outside = TRUE, 
            legend.outside.position = "right")


tmap_arrange(torture_map, rep_map, disappearance_map)
```

The next step in my analysis was to map out socioeconomic factors by
governante. To do so I joined the socioeconomic data with the Egypt
shape file and used tmap to create another interactive map. The
socioeconomic data was already grouped by governate's English names, but
those names needed to be standardized with the Egypt shapefile. I chose
3 factors to map- Poverty, Life Expectancy, and Education.

```{r socioeconomic}
# interactive tmap

tmap_mode('view')

{egypt %>% 
    left_join(socioeconomic %>% 
                mutate(Region = case_when(
                  Region == "Assuit" ~ 'Assiut',
                  Region == "Kafr El-Sheikh" ~ "Kafr El-Shikh",
                  Region == "Kalyubia"~ "Kalyoubia",
                  Region == "Menya" ~ "Menia",
                  Region == "Souhag" ~ "Suhag",
                  TRUE ~ Region)) %>% 
                select('Region', 
                       'Poverty (IWI<50)'), 
              by = c("ADM1_EN" = "Region")) %>% 
    
    tm_shape(name = 'Poverty') +
    tm_polygons(col = 'Poverty (IWI<50)') +
    
    egypt %>% 
    left_join(socioeconomic %>% 
                mutate(Region = case_when(
                  Region == "Assuit" ~ 'Assiut',
                  Region == "Kafr El-Sheikh" ~ "Kafr El-Shikh",
                  Region == "Kalyubia"~ "Kalyoubia",
                  Region == "Menya" ~ "Menia",
                  Region == "Souhag" ~ "Suhag",
                  TRUE ~ Region)) %>% 
                select('Region', 
                       'Mean Years Schooling'), 
              by = c("ADM1_EN" = "Region")) %>% 
    
    tm_shape(name = 'Mean Years Schooling') +
    tm_polygons(col = 'Mean Years Schooling') +
    
    tm_basemap('Esri.WorldTopoMap') +
    tm_basemap('Esri.WorldStreetMap') +
    
    egypt %>% 
    left_join(socioeconomic %>% 
                mutate(Region = case_when(
                  Region == "Assuit" ~ 'Assiut',
                  Region == "Kafr El-Sheikh" ~ "Kafr El-Shikh",
                  Region == "Kalyubia"~ "Kalyoubia",
                  Region == "Menya" ~ "Menia",
                  Region == "Souhag" ~ "Suhag",
                  TRUE ~ Region)) %>% 
                select('Region',
                       'Life expectancy'), 
              by = c("ADM1_EN" = "Region")) %>% 
    
    tm_shape(name = 'Life expectancy') +
    tm_polygons(col = 'Life expectancy')} %>% 
  
  tmap_leaflet() %>%
  leaflet::hideGroup('Life expectancy') %>% 
  leaflet::hideGroup('Mean Years Schooling')
```

I hypothesized that socioeconomic factors would be a significant
influencing factor on both initial detention and resulting abuses. To
test this theory I calculated the correlation between various abuses and
between those abuses and the socioeconomic factors that I had mapped. To
do so I had to create an additional column mapping any reported torture
onto a binary indicator of whether or not torture was reported.

```{r correlation}
# create a column for presence of torture and correlate with other abuses

abuse_correlation <- 
  cor(joined_data %>% 
        as_tibble() %>% 
        mutate(torture = if_else(
          is.na(beating) == FALSE | 
            is.na(electrocution) == FALSE |
            is.na(threaten_family) == FALSE |
            is.na(hanging) == FALSE,
          1,
          0)) %>% 
        mutate(
          lawyer_present = case_when(
            lawyer_present == Present ~ 1,
            lawyer_present != Present & 
              is.na(lawyer_present) == FALSE ~ 0)) %>% 
        select("Disappeared" = disappeared, 
               "Torture" = torture, 
               'Legal Rep' = lawyer_present) %>% 
        drop_na())  

# plot correlation

corrplot(abuse_correlation, 
         type="upper", 
         order="hclust",
         title = "Correlation Between Abuses",
         tl.cex = 0.8,
         mar=c(0,0,1,0))

# correlation between abuses and socioeconomic factors

socioeconomic_corr <- 
  cor(joined_data %>% 
        as_tibble() %>% 
        mutate(torture = if_else(
          is.na(beating) == FALSE | 
            is.na(electrocution) == FALSE |
            is.na(threaten_family) == FALSE |
            is.na(hanging) == FALSE,
          1,
          0)) %>% 
        mutate(
          lawyer_present = case_when(
            lawyer_present == Present ~ 1,
            lawyer_present != Present & 
              is.na(lawyer_present) == FALSE ~ 0)) %>% 
        select(ADM1_EN, 
               disappeared, 
               torture, 
               lawyer_present) %>% 
        drop_na() %>%
        group_by(ADM1_EN) %>% 
        summarise(Torture = sum(torture), 
                  "Disappeared" = sum(disappeared), 
                  "Legal Rep" = sum(lawyer_present)) %>% 
        left_join(socioeconomic %>% 
                    mutate(Region = case_when(
                      Region == "Assuit" ~ 'Assiut',
                      Region == "Kafr El-Sheikh" ~ "Kafr El-Shikh",
                      Region == "Kalyubia"~ "Kalyoubia",
                      Region == "Menya" ~ "Menia",
                      Region == "Souhag" ~ "Suhag",
                      TRUE ~ Region)) %>% 
                    select('Region', "Poverty" = 'Poverty (IWI<50)', 
                           "Education" = 'Mean Years Schooling'), 
                  by = c("ADM1_EN" = "Region")) %>% 
        select(-ADM1_EN) %>% 
        drop_na())  

# plot correlation 

corrplot(socioeconomic_corr, 
         type="upper", 
         order="hclust",
         title = "Socioeconomic Correlation",
         tl.cex = 0.8,
         mar=c(0,0,1,0))
```

# Shiny App

To display all of the analysis above, and to provide relevant context
and interpretation, I created a shiny app. That app is included in this
project file as well as in the github repository.

The app contains a home tab providing general context on politically
motivated arbitrary detention in Egypt. That tab is then followed by
additional pages showing the interactive maps, the static maps, the
static plots of abuse, the correlation plots, and finally summary tables
displaying statistics of abuse by governate. Each page also includes
additional context and an interpretation of the analysis
